---
title: "Counts and Cumulative Sums by Group"
description: |
  A short description of the post.
author:
  - name: Jeremy Allen
    url: https://jeremydata.com
date: 03-13-2021
output:
  distill::distill_article:
    self_contained: false
categories:
    - data.table
    - time series
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Problem:

**In a time series, how do I add a counter and cumulative sum that resets every 30 minutes?**  

I keep forgetting about R's `seq_len()` function. I'm writing about it here, so when I inevitably google for it I can find it here. Let's see how it works.

We will be using [**data.table**](https://rdatatable.gitlab.io/data.table/index.html) for this example. 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(data.table)
library(lubridate)


#---- fake data ----

times <- seq.POSIXt(
 from = as.POSIXct("2021-03-12 08:00:00"),
 to =   as.POSIXct("2021-03-12 12:00:00"),
 by = "sec"
)

accounts <- seq(1111, 9999, by = 1000)

amounts <- seq(1, 5000, by = 1)

set.seed(865)

dt <- data.table(
 datetime = sample(times, 1500, replace = TRUE),
 account = sample(accounts, 1500, replace = TRUE),
 amount = sample(amounts, 1500, replace = TRUE)
)

# order by account and datetime
setkeyv(dt, c("account", "datetime"))

```

## n = seq_len(.N) By Group Row Number in data.table  

After making our fake time series data, we can uses `seq_len(.N)` along with the `by` argument in data.table to add a counter that resets for each 30-minute group. Notice the transformation of a variable within the `by =` of the data.table.

**Cumulative sum by 30-minute periods**  


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, layout="l-body-outset"}

#---- counts and sums ----

# count and sum by half hour using seq_len() to show 
# each observation's position within its 30-minute group
dt[,
   `:=`(n_by_30_mins = seq_len(.N), cumulative_amount = cumsum(amount)) ,
   by = .(account, floor_date(datetime, "30 minutes"))]

head(dt, 50)
```
