---
title: "By Row in Base R, Tidyverse, and data.table"
description: |
  Which rows have more 1s than 0s?.
author:
  - name: Jeremy Allen
    url: https://jeremydata.com
date: 03-12-2021
preview: solution.png
output:
  distill::distill_article:
    self_contained: false
categories:
  - rowwise
  - tidyverse
  - data.table
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## PROBLEM: Add a column that indicates if a row has more 1s than 0s, use 1 if true, 0 if false, but only consider specific columns and allow for NAs.

This question was asked on Twitter, and I want to elaborate on the solutions here.  

We know we need to include

- a column selection constraint

- properly handle NAs

- answers in a new column

- 1 for TRUE and 0 for FALSE  

First, let's make sample data that includes a row with NA.  

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# PROBLEM: add a column that indicates if a row has more 1s than 0s,
# use 1 if true, 0 if false, but only consider certain columns and
# allow for NAs.

# sample data
df <- data.frame(
  a = as.integer(c(1,0,1,0,0)),
  b = as.integer(c(1,0,1,0,0)),
  c = as.integer(c(1,0,1,1,0)),
  d = as.integer(c(1,1,0,9,0)),
  e = as.integer(c(1,1,0,1,0)),
  f = as.integer(c(1,1,NA,0,0)),
  g = as.integer(c(1,1,1,0,0))
)

```

Let's try a base R solution first.  Base R has a function `rowMeans()` that could be helpful.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

#---- base R solution ----

# base R rowMeans() says yes to row 4, but row 4 does not have 
# more 1s than 0s, and it will fail on any non-numeric columns
df$more_1s <- ifelse(rowMeans(df, na.rm = T) > .5,1,0)

df

```

However, if use the mean of each row, instead of literally just counting 1s and 0s, then we can be fooled by rows with larger numbers, like row 4 above. So, let's make a function that only counts 1s and 0s, which is the problem we were asked to solve.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# function to count only ones and zeros and report TRUE or
# FALSE if more 1s, and convert the logical to integer
is_more_1s <- function(x) {
  as.integer(
    sum(x == 1, na.rm = T) > sum(x == 0, na.rm = T)
  )
}

# this gets row 4 correct
df$more_1s <- apply(df, 1, is_more_1s)

df

```

This function only counts 1s and 0s, ignoring the 9 in row 4 and therefore giving us the correct answer in our new column. The function also returns a 1 if `TRUE` and a 0 if `FALSE`.  

## library(tidyverse)

Let's try a tidyverse solution that uses `rowwise()` instead of `apply()`

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

#---- tidyverse solution ----

# sample data
df <- data.frame(
  a = as.integer(c(1,0,1,0,0)),
  b = as.integer(c(1,0,1,0,0)),
  c = as.integer(c(1,0,1,1,0)),
  d = as.integer(c(1,1,0,9,0)),
  e = as.integer(c(1,1,0,1,0)),
  f = as.integer(c(1,1,NA,0,0)),
  g = as.integer(c(1,1,1,0,0))
)

library(tidyverse)

tb <- as_tibble(df)
tb %>% 
  rowwise() %>% 
  mutate(more_1s = is_more_1s(c_across(a:g))) # only a through g

```

Here, `rowwise()` makes sure we are counting across rows, and `c_across()` lets us constrain which columns we consider, which is part of the problem we were asked to solve.  

## library(data.table)

Using only data.table we don't have `rowwise()` from dplyr, so we use base R's `apply()` again. We use `.SD` and `.SDcols` to specify which columns we want to be constrained to.  

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

#---- data.table solution ----

# sample data
df <- data.frame(
  a = as.integer(c(1,0,1,0,0)),
  b = as.integer(c(1,0,1,0,0)),
  c = as.integer(c(1,0,1,1,0)),
  d = as.integer(c(1,1,0,9,0)),
  e = as.integer(c(1,1,0,1,0)),
  f = as.integer(c(1,1,NA,0,0)),
  g = as.integer(c(1,1,1,0,0))
)

library(data.table)

dt <- as.data.table(df)
my_cols <- letters[1:7] # only a through g
dt[, more_1s := apply(.SD, 1, is_more_1s), .SDcols = my_cols]

dt

```

