---
title: "Order Months"
description: |
  Produce a vector of month names ending with the current month.
author:
  - name: Jeremy Allen
    url: https://jeremydata.com
date: 05-01-2020
preview: order_months.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The last 12 months with order_months()

I frequently have to make tables or plots of events that "happened in the last 12 months." I like to generate a character vector of month names for the last 12 months which I can then use for ordered factor levels, factor labels, column names, or x-axis labels.

This helper function will produce a character vector of names of the last 12 months ending either at the current month or a given month. The current month or given month will be the last element in the vector, with the preceding 11 months before it.

If no month number is given for x, the current month is used. Given months must be given as a number. Month names can be returned as abbreviated or full names. Set `label = "abb"` or `label = "names"`. X must be 1-12.

## Remember, the current or given month is always at the end of the vector!

```{r function, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

order_months <- function(x = NULL, label = "abb") {
  
  # This function takes a given month number or the current month
  # number and returns a character vector of the last 12 months,
  # including current month. For example, if it is now February: 
  # "Mar" "Apr" "May" "Jun" "Jul" "Aug" "Sep" "Oct" "Nov" "Dec" "Jan" "Feb"
  # is returned with the current month at the end.
  # This makes a nice x axis if you need to plot something for
  # the "last 12 months".
  # if x is NULL the current month is taken from Sys.Date
  # if x is given, x must be 1-12
  # if label is "abb", month names are abbreviated
  # if label is "names", month names are full
  
  # stop if x is out of bounds
  if(!is.null(x))
    if(x > 12 | x < 1) stop("x must be 1-12")
  
  # sets of named integers using abbreviated and full month names
  months_abb <- setNames(1:12, month.abb)
  months_names <- setNames(1:12, month.name)
  
  # current month number
  m <- as.POSIXlt(Sys.Date())$mon + 1
  
  # get a value for x
  if(is.null(x)) x <- m + 1 else x <- x + 1
  
  # a and b components for main if
  # if x is 1 or 12
  a <- 1:12
  # if x is 2 through 12
  b <- c(
    x:12,
    1:(x-1)
    )
  
  # main if
  if(x == 1 | x == 13) new_order <- a else new_order <- b

  # use new_order to set desired order of months
  if(label == "abb") my_month_order <- names(months_abb[new_order])
  if(label == "names") my_month_order <- names(months_names[new_order])
  
  my_month_order
}
```

## Typical usage

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
order_months()
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
order_months(label = "names")
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
order_months(x = 5, label = "abb")
```
