---
title: "Find Index of First Instance"
description: |
  Find one thing with another thing. We'll speed test various data.table and tidyverse methods for finding the position of the first match and use that to index another column.
author:
  - name: Jeremy Allen
    url: https://jeremydata.com
date: 05-11-2020
preview: index_benchmark.png
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Yes, another data.table and tidyverse speed test, but this is more than that! I swear. This is real code of me working through a specific real-world issue.

__The problem: Get the position number of the first instance of a thing in one column, and use that number to pick a thing from another column, in my case, returning the first date on which a specified number of cases occurred.__

Load packages and make some fake data.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(data.table)
library(dplyr)
library(purrr)

# lots of dates
date = seq.Date(from = as.Date("1900-01-01"),
                to = as.Date("2900-12-31"),
                by = "day")

# lots of cases
cases = c(1:length(date)-1)

# make a dataframe
df <- data.frame(date = date, cases = cases)

```

<br>

Let's use `which.max()` to get the date on which the first instance of 10,0000 cases occurred.

```{r which_max, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Get the position of the first instance of 10,000 in the cases col,
# and use that number to index the date col, returning the first date
# on which 10,000 cases occurred.
dt <- as.data.table(df) # convert to data.table first
dt[, date[which.max(cases >= 10000)]]
# this only works because 10000 is a vlaue that can be found in that column.
```

<br>

However, `which.max()` returns 1 when it fails, thus indexing our first date, which we do not want because there are no days with 400,000 or more cases.

__We need NA returned when we don't find an instance of the value we are looking for.__

```{r which_max_fail, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# which.max returns 1 when it fails, thus indexing
# our first date, which we do not want because there
# are no days with 400,000 or more cases. We expect NA.
dt[, date[which.max(cases >= 400000)]]

```

<br>

Let's test many methods. We want to speed test them at the end, so I'm putting each method inside a function because it's easier to add them as functions in the speed test once we get there.

```{r functions, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# which.max(), does NOT return NA when it fails. Bad.
dt_which_max_method <- function() {
  dt <- as.data.table(df)
  dt[, date[which.max(cases >= 400000)]]
} 

# match(true, x) will return NA when it fails, which
# is what we want so that we don't get a date returned
# when there are no days with 400,000 or more cases
dt_match_true_method <- function() {
  dt <- as.data.table(df)
  dt[, date[match(TRUE, cases >= 100000)]]
} 

# which()[1], test them all and return the first one, also returns NA
dt_which_first_method <- function() {
  dt <- as.data.table(df)
  dt[, date[which(cases >= 400000)[1L]]]  
}

# use base R's Position function, also returns NA
dt_position_method <- function() {
  dt <- as.data.table(df)
  dt[, date[Position(function(x) x >= 400000, cases)]]
}

# Tidyverse's purrr::detect_index(), returns 'Date of length 0'
tv_purrr_method <- function() {
  tb <- tibble::as_tibble(df)
  tb %>%
    slice(purrr::detect_index(cases, ~.x >= 400000)) %>% 
    pull(date)
}

# Tidyverse mixed with the base R's match function
tv_match_method <- function() {
  tb <- tibble::as_tibble(df)
  tb %>%
    slice(match(TRUE, cases >= 100000)) %>% 
    pull(date)
}

```

<br>

Get each function into microbenchmark and test each one 100 times.

```{r speed_test, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, layout="l-body-page"}

#--- Speed test them each 100 times

microbenchmark::microbenchmark(
  dt_which_max_method(),
  dt_match_true_method(),
  dt_which_first_method(),
  dt_position_method(),
  tv_purrr_method(),
  tv_match_method(),
  times = 100L
)

```

<br>

The vectorized methods, such as `match()` used on either a data.table or a tidyverse tibble are clear winners over the base `Position()` and purrr `detect_index()` functions.